---
title: "SLU DSS Spring 2020"
subtitle: "Statistics in R: Session #3"
author: "Cort W. Rudolph"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: ioslides_presentation
widescreen: true
smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = FALSE,
  message = FALSE,
  warning = FALSE
  )
options(scipen = 99, digits = 3)
```

# Statistics in R|Session #2: Modeling Continuous Predictors

## Needed packages
- Let's load all the packages needed for this week. 
- `ipak` codeblock available in the `RMarkdown` file:

<!-- To install & load the required packages for this lecture, run this codeblock -->
```{r echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# ipak <- function(pkg){
#     new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
#     if (length(new.pkg)) 
#         install.packages(new.pkg, dependencies = TRUE)
#     sapply(pkg, require, character.only = TRUE)
# }
# 
# # usage
# packages <- c("tidyverse", 
#               "gapminder", 
#               "broom",
#               "knitr,"
#               "kableExtra",
#               "emmeans")
# ipak(packages)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(gapminder)
library(broom)
library(knitr)
library(kableExtra)
library(emmeans)
```

# Modeling Categorical Predictors

## Modeling Categorical Predictors
Motivating research question:

- "Are there significant differences in life expectancy, on average, between the five continents of the world: Africa, the Americas, Asia, Europe, and Oceania?"
    
## Modeling Categorical Predictors
- To answer this questions, we'll use the `gapminder` dataset in the `gapminder` package. 
- This dataset has international development statistics such as life expectancy, GDP per capita, and population by country ($n$ = 142) for 5-year intervals between 1952 and 2007. 

## Modeling Categorical Predictors
- We'll use this data for linear regression again, but note that our explanatory variable $x$ is now categorical, and not numerical like when we covered simple linear regression. More precisely, we have:

    1. A numerical outcome variable $y$. In this case, life expectancy.
    1. A single categorical explanatory variable $x$, In this case, the continent the country is part of. 

## Modeling Categorical Predictors
- Let's load the `gapminder` data and `filter()` for only observations in 2007. 
- Then `select()` only the variables we'll need 
- Lastly, we save this in a data frame with name `gapminder2007`:

```{r, warning=FALSE, message=FALSE}
gapminder2007 <- gapminder %>%
  filter(year == 2007) %>% 
  select(country, continent, lifeExp)
```

## Modeling Categorical Predictors

First, let's take a look at the structure of our data:

```{r  echo=FALSE}
gapminder2007 %>%
  sample_n(5) %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

## Modeling Categorical Predictors

Next, let's create a table of descriptive statistics (i.e., means and SDs) by continent.

These are the means (and the associated variability around them) that our statistical model is "modeling"

```{r, echo=FALSE}
lifeExp_by_continent <- gapminder2007 %>%
  group_by(continent) %>%
  summarize(mean = mean(lifeExp), sd = sd(lifeExp)) 

lifeExp_by_continent %>%
  kable() %>% 
  kable_styling(full_width = FALSE)
```

## Modeling Categorical Predictors
- Now, let's start making comparisons of life expectancy *between* continents. 
- Let's use Africa as a *baseline for comparison*. 
- Why Africa? Two reasons: 1) It has the lowest life expectancy, 2) It comes first alphabetically (and `R` defaults to alphabetical order to determine the baseline for comparison)

## Modeling Categorical Predictors
- Before we fit our regression model, let's create a summary table that reports the difference in mean life expectancy *relative* to Africa's mean life expectancy:

```{r echo=FALSE}
mean_africa <- lifeExp_by_continent %>%
  filter(continent == "Africa") %>%
  pull(mean)

gapminder2007 %>%
  group_by(continent) %>%
  summarize(mean = mean(lifeExp)) %>%
  mutate(`mean vs Africa` = mean - mean_africa) %>% 
  kable() %>% 
  kable_styling(full_width = F)
```

## Modeling Categorical Predictors
- Now, let's use the `tidy()` function to get the *regression table* for the `gapminder2007` regression analysis: 

## Modeling Categorical Predictors
```{r, eval=FALSE}
lm(lifeExp ~ continent, data = gapminder2007) %>% 
  tidy(conf.int=TRUE) %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

## Modeling Categorical Predictors

You might also be interested in model performance (e.g., "Does `continent` account for an *appreciable* amount of the variability in `lifeExp`)

To do so, we can ask for an `anova` table, which gives us an omnibus test of $R^2$ (and answers our question, posed above)

```{r}
lm(lifeExp ~ continent, data = gapminder2007) %>% 
  anova() %>%
  tidy() %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

## Modeling Categorical Predictors

As before, we can also get the corresponding $R^2$
```{r}
lm(lifeExp ~ continent, data = gapminder2007) %>% 
  summary() %>% 
  .$r.squared
```

## Modeling Categorical Predictors

We can now ask another more nuanced question: "Which `continent`s differ from one-anther in terms of `lifeExp`?"

To do so, we will estimate marginal means using `emmmeans::emmmeans`. 

`emmmeans` work exactly like `ggeffects` used last time we met (i.e., for making predictions out of `lm` objects), but has extended flexibility for working with categorical predictors.

```{r}
lm(lifeExp ~ continent, data = gapminder2007) %>% 
  emmeans(. , ~continent) %>% 
  pairs() %>%
  # confint() %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

## Modeling Categorical Predictors

Visualizing mean differences:

- The blue bars are 95% confidence intervals for the EMMs, and the red arrows are for the comparisons among them. 
- If an arrow from one mean overlaps an arrow from another group, the difference is not “significant."
- Based on the `adjust` setting (which defaults to "tukey") and the value of `alpha` (which defaults to 0.05).

```{r}
lm(lifeExp ~ continent, data = gapminder2007) %>% 
  emmeans(. , ~continent) %>% 
  plot(comparisons=TRUE)
```




# Your Turn!

## CDC Data
The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone survey of 350,000 people in the United States conducted by the CDC

Here, we will focus on a random sample of 20,000 people from the BRFSS survey conducted in 2000. While there are over 200 variables in this data set, for now, we will work with a small subset (i.e., just respondents height & weight):

First, we load the data:
```{r}
source("http://www.openintro.org/stat/data/cdc.R")
```

Then, select a random sample of $n = 100$ respondents `weight`, `wtdesire`, and `genhlth`, and create a new variable `weight_goal` (`weight_goal` < 1 == lose weight, `weight_goal` > 1 == gain weight)
```{r}
set.seed(1818)
cdc_subset<-cdc %>%
  sample_n(100) %>%
  select(weight, wtdesire, genhlth) %>%
  mutate(weight_goal = wtdesire - weight)
```

# Table to demonstrate data structure
```{r}
cdc_subset %>%
  sample_n(5) %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

# Estimate a model where `weight_goal` is regressed onto `genhlth`
```{r, eval=FALSE}
lm(weight_goal ~ genhlth, data = cdc_subset) %>% 
  tidy(conf.int=TRUE) %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

# Which levels of `genhlth` differ from one-another on `weight_goal`?
```{r}
lm(weight_goal ~ genhlth, data = cdc_subset) %>% 
  emmeans(. , ~genhlth) %>% 
  pairs() %>%
  # confint() %>%
  knitr::kable() %>% 
  kable_styling(full_width = F)
```

# Plot these comparisons
```{r}
lm(weight_goal ~ genhlth, data = cdc_subset) %>% 
  emmeans(. , ~genhlth) %>% 
  plot(comparisons=TRUE)
```
